---
title: "Figure 3 manuscript panels (participant stool transfer)"
output: html_document
---

Manuscript-facing copy of the participant-stool transfer analyses in `figure_3.Rmd`. The original notebook is unchanged.

In the submitted figure, the mimic-10 tumor-volume curves are panel A and the mixed-model heatmap is panel D. The antibiotics timeline and mouse icons on panel A were added outside R. Panel B is drawn by `mimic_gavaged-sample_barplots.Rmd`. Panel C is drawn by `MetaG_16S_PCoA_LN.Rmd`.

Knit from `manuscript/scripts/`.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(ggtext)
library(viridis)
```

```{r}
load("../data/figure-data/tvol_m10.rda")
load("../data/figure-data/boxdat_mimic.rda")
load("../data/figure-data/longmods_mimic-heatmap.rda")
```

```{r tumor-volume}
mimic10.means %>%
  mutate(Gavage = if_else(grepl("v1", treatment.group), "Pre-BRB", "Post-BRB"),
         Treatment = if_else(grepl("Anti", treatment.group), "Anti-PD1", "IgG"),
         Gavage = fct_relevel(Gavage, c("Pre-BRB"))) %>%
  ggplot(aes(x = days.from.injection, y = mean)) +
  geom_line(aes(group = treatment.group, color = Gavage, linetype = Treatment),
            position = position_dodge(width = .5), lwd = 1) +
  geom_point(aes(color = Gavage), position = position_dodge(width = .5)) +
  scale_color_viridis_d(direction = -1) +
  geom_errorbar(aes(ymin = lwr, ymax = upr, color = Gavage),
                width = 1, position = position_dodge(width = .5), lwd = 1) +
  labs(x = "Days from Tumor Injection",
       y = expression("Tumor volume ("~mm^3~")")) +
  theme_minimal(base_size = 20) +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        plot.background = element_blank(),
        axis.line = element_line()) +
  scale_y_continuous(position = "right")

ggsave("../figures/panels/fig3A_mimic10-tumor-volume.svg", height = 3.5, width = 9)
```

```{r lme-heatmap}
mimic.longmods %>%
  filter(term.adj %in% c("Anti-PD1", "Gavage")) %>%
  mutate(Combined_ID2 = gsub(".*\\(([^)]+)\\).*", "\\1", `Combined ID`)) %>%
  mutate(Combined_ID2 = case_when(
    Combined_ID2 == "DL081" ~ "Control-NR",
    Combined_ID2 == "DL017" ~ "Control-R",
    Combined_ID2 == "Control-NR" ~ "Control-R",
    Combined_ID2 == "Control-R" ~ "Control-NR",
    TRUE ~ Combined_ID2
  )) %>%
  ggplot(aes(x = Combined_ID2, y = term.adj, fill = estimate, alpha = sig.flag)) +
  geom_tile(color = "black") +
  scale_fill_viridis(option = "magma") +
  coord_fixed() +
  guides(alpha = "none") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 20, hjust = 1)) +
  labs(x = "Test Microbe (Sample ID)", y = "")

ggsave("../figures/panels/fig3D_lme-heatmap.svg", height = 4, width = 8)
write_csv(mimic.longmods, "../tables/Supp-Table-S4_model-summary-values.csv")
```
