---
title: "Figure 2 manuscript panels (BE WELL microbiome in this paper)"
output: html_document
---

Manuscript-facing copy of `figure_2.Rmd`. The original notebook is unchanged. Knit this file from `manuscript/scripts/`.

The human trial itself, including enrollment, nectar, urine polyphenols, and cytokines, is in the companion repository [spakowiczlab/bewell](https://github.com/spakowiczlab/bewell). This notebook is only the microbiome analyses reported in this mouse-model paper.

Panel letters follow the submitted figure. Figure 2E is the responder-signature ordination. Figure 2F is the change in distance to the responder and non-responder centroids. `figure_2.Rmd` saved those two plots under the opposite letters. Figure 2A is the trial schema and is not drawn here. The bar chart in the submitted Figure 2D is `manuscript/figures/2D.svg`; the effect-size chunk below is that analysis.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(ggrepel)
library(ggforce)
library(ggtext)
library(ggplot2)
library(cowplot)
library(gridExtra)
library(vegan)
```

```{r}
load("../data/figure-data/stackedbar_bewell.rda")
load("../data/figure-data/model_stats_BW.rda")
load("../data/figure-data/bewell-respondersig-pca.rda")
load("../data/figure-data/bewell-respondersig-dist.rda")
```

```{r fig2B-stacked-bar}
stackdatBW %>%
  mutate(timepoint = str_replace(timepoint, "-", " "),
         timepoint = factor(timepoint, levels = c("Pre Placebo", "Post Placebo", "Pre BRB", "Post BRB"))) %>%
  ggplot(aes(x = sample, y = RelAbun, fill = Phylum)) +
  facet_row(vars(timepoint), scales = "free_x", space = "free", strip.position = "bottom") +
  geom_bar(position = "fill", stat = "identity") +
  labs(x = "", y = "") +
  scale_fill_viridis_d(option = "plasma") +
  theme_minimal(base_size = 9) +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        axis.text.y = element_blank(),
        panel.grid.major = element_blank(),
        strip.text.x = element_text(size = 19),
        legend.text = element_text(size = 20),
        legend.title = element_text(size = 24))

ggsave("../figures/panels/fig2B_phylum-stacked-bars.svg", width = 10, height = 4)
```

```{r fig2C-volcano}
volplot.in <- mic_fc_BW %>%
  ungroup() %>%
  right_join(model_res_BW, by = "term") %>%
  mutate(
    tlev = str_sub(term, 1, 1),
    tlab = case_when(
      term == "s__Roseburia_sp_CAG_309" ~ "Roseburia sp. CAG309",
      p.value < 0.05 ~ gsub("_", " ", gsub("\\w__", "", term)),
      TRUE ~ NA_character_
    )
  )

volplot.in %>%
  ggplot(aes(x = l2fc, y = -log(p.value), color = tlev, label = tlab)) +
  geom_point() +
  geom_text_repel(show.legend = FALSE, nudge_x = -0.6, size = 6.75, fontface = "italic") +
  scale_color_brewer(palette = "Dark2", type = "qual",
                     breaks = c("k", "p", "c", "o", "f", "g", "s"),
                     labels = c("Kingdom", "Phylum", "Class", "Order", "Family", "Genus", "Species"),
                     name = "Taxonomic\nRank") +
  labs(x = "Log2FoldChange", y = "-log(p-value)") +
  theme_bw(base_size = 18) +
  theme(legend.position = c(0.97, 0.05),
        legend.justification = c("right", "bottom"),
        legend.box.just = "right",
        legend.margin = margin(3, 3, 3, 3),
        legend.box.background = element_rect(color = "black", linewidth = 0.25),
        legend.box.margin = margin(2, 2, 2, 2),
        legend.text = element_text(size = 16),
        legend.title = element_text(size = 18, hjust = 0.5))

ggsave("../figures/panels/fig2C_bewell-taxon-volcano.svg", height = 5, width = 7)
write_csv(model_res_BW, "../tables/Supp-Table-S2_bewell-volcano.csv")
```

```{r fig2D-effect-size}
model_res_BW %>%
  mutate(padj = p.adjust(p.value, method = "fdr")) %>%
  filter(p.value < 0.08) %>%
  separate(term, into = c("TaxaLevel", "term"), sep = "__") %>%
  mutate(term = gsub("_", " ", term),
         term = paste0("*", term, "*"),
         term = gsub("\\*Lachnospiraceae unclassified\\*", "Unclassified Lachnospiraceae", term),
         term = fct_reorder(term, estimate)) %>%
  ggplot(aes(x = estimate, y = term, fill = TaxaLevel)) +
  geom_col() +
  labs(x = "Effect Size", y = "") +
  scale_fill_brewer(palette = "Dark2", breaks = c("f", "g", "s"),
                    name = "Taxonomic\nRank",
                    labels = c("Family", "Genus", "Species")) +
  theme_bw(base_size = 10) +
  theme(axis.text.y = element_markdown(),
        legend.title = element_text(hjust = 0.5))

ggsave("../figures/panels/fig2D_bewell-effect-sizes.svg", height = 3, width = 4)
```

```{r fig2F-centroid-change}
score <- bewellsig.dist %>%
  filter(CentroidGroup == "Responder" | CentroidGroup == "Non_Responder") %>%
  mutate(time = sub("\\-.*", "", Group),
         Group = sub(".*-", "", Group),
         Sample = sub("\\-.*", "", sample)) %>%
  dplyr::select(Sample, Group, time, CentroidGroup, CentroidDistance) %>%
  pivot_wider(names_from = time, values_from = CentroidDistance) %>%
  mutate(Improvement = Pre - Post)

score$Group <- factor(score$Group, levels = c("Placebo", "BRB"))

score %>%
  ggplot(aes(y = Improvement, x = Group, fill = CentroidGroup, color = Group,
             shape = Group, alpha = 0.5)) +
  scale_fill_manual(values = c("gray", "darkolivegreen3")) +
  scale_color_manual(values = c("dodgerblue3", "darkorchid4")) +
  scale_shape_manual(values = c(17, 18)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(aes(size = Group), show.legend = FALSE) +
  scale_size_manual(values = c(2, 3)) +
  theme_bw() +
  ylab("Change in Distance to Centroid (Pre-Post)") +
  xlab("") +
  guides(color = "none", shape = "none", alpha = "none") +
  coord_cartesian(xlim = c(-0.1, 2.5))

ggsave("../figures/panels/fig2F_change-in-distance-to-centroid.svg", width = 7, height = 5)
```

```{r fig2E-responder-pca}
data <- bewellsig.pca[[1]]
expl.var <- bewellsig.pca[[2]]
hull <- data$hull

pMain <- ggplot(data = data, aes(x = PC1, y = PC2, color = response, shape = response)) +
  geom_point(aes(size = response)) +
  xlab(paste0("PC1: ", round(as.numeric(expl.var[1]) * 100), "% expl.var")) +
  ylab(paste0("PC2: ", round(as.numeric(expl.var[2]) * 100), "% expl.var")) +
  scale_shape_manual(values = c(1, 1, 15, 17, 16, 18)) +
  scale_size_manual(values = c(1, 1, 2, 2, 2, 3)) +
  scale_color_manual(values = c("darkolivegreen3", "#999999", "firebrick3", "dodgerblue3",
                                "darkgoldenrod1", "darkorchid4")) +
  stat_ellipse(aes(group = hull, color = hull, linetype = hull), size = 0.2, show.legend = FALSE) +
  scale_linetype_manual(values = c("solid", "solid")) +
  theme_bw() +
  theme(legend.position = "none")

xlim.update <- layer_scales(pMain)$x$get_limits()
ylim.update <- layer_scales(pMain)$y$get_limits()

pTop <- ggplot(data = data, aes(x = PC1, fill = response)) +
  geom_density(size = 0.2, alpha = 0.5) +
  ylab("Density") +
  theme(axis.title.x = element_blank(),
        axis.title.y = element_text(size = rel(0.8)),
        axis.line = element_blank(),
        axis.text = element_blank(),
        axis.ticks = element_blank(),
        panel.background = element_blank(),
        legend.position = "none") +
  scale_fill_manual(values = c("darkolivegreen3", "#999999", "firebrick3", "dodgerblue3",
                               "darkgoldenrod1", "darkorchid4")) +
  scale_x_continuous(limits = xlim.update)

pRight <- ggplot(data = data, aes(x = PC2, fill = response)) +
  geom_density(size = 0.2, alpha = 0.5) +
  coord_flip() +
  ylab("Density") +
  theme(axis.title.x = element_text(size = rel(0.8)),
        axis.title.y = element_blank(),
        axis.line = element_blank(),
        axis.text = element_blank(),
        axis.ticks = element_blank(),
        panel.background = element_blank(),
        legend.position = "none") +
  scale_fill_manual(values = c("darkolivegreen3", "#999999", "firebrick3", "dodgerblue3",
                               "darkgoldenrod1", "darkorchid4")) +
  scale_x_continuous(limits = ylim.update)

legend <- get_legend(
  ggplot(data = data, aes(x = PC1, y = PC2, color = response, shape = response)) +
    geom_point(aes(size = response)) +
    scale_shape_manual(values = c(1, 1, 15, 17, 16, 18)) +
    scale_size_manual(values = c(1, 1, 2, 2, 2, 3)) +
    scale_color_manual(values = c("darkolivegreen3", "#999999", "firebrick3", "dodgerblue3",
                                  "darkgoldenrod1", "darkorchid4")) +
    theme(legend.position = "right", legend.box = "horizontal",
          legend.direction = "vertical",
          legend.key.height = unit(0.2, "cm"),
          legend.key.width = unit(0.1, "cm"),
          legend.title = element_blank(),
          legend.spacing.x = unit(0.1, "cm"),
          legend.spacing.y = unit(0.1, "cm"),
          legend.text = element_text(size = rel(0.7)))
)

g <- grid.arrange(pTop, legend, pMain, pRight,
                  ncol = 2, nrow = 2, widths = c(3, 1), heights = c(1, 3))

ggsave("../figures/panels/fig2E_responder-signature-pca.svg", g, height = 5, width = 7)
```
