---
title: "Short-chain fatty acids shown with the Blautia metabolomics"
output: html_document
---

Manuscript-facing copy of the short-chain fatty acid panel in `metabolomics.Rmd`. The original notebook is unchanged and still writes the individual indole and blood metabolite plots.

In the submitted manuscript, butyrate, heptanoate, and valerate are Figure 5C. The ten-acid panel written by `metabolomics.Rmd` to `SCFA-plot.svg` is Supplementary Figure S1. Knit from `manuscript/scripts/`.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(readxl)
```

```{r}
mimic_24_SCFA_Data_Summary <- read_excel(
  "../../exploratory/data/metabolomics/mimic 24_SCFA_Data_Summary.xlsx",
  sheet = "Normalization", n_max = 24
)
mimic_24_SCFA_Data_Summary[is.na(mimic_24_SCFA_Data_Summary)] <- 0

normalized_cleaned_data <- mimic_24_SCFA_Data_Summary %>%
  mutate(across(c(`acetic acid`, `propionic acid`, `isobutyric acid`, `butyric acid`,
                  `2-methylbutyric acid`, `isovaleric acid`, `valeric acid`,
                  `isocaproic acid`, `caproic acid`, `heptanoic acid`),
                ~ . / `feces weight for SCFA (mg)`,
                .names = "norm_{col}")) %>%
  mutate(mouse = gsub("neg-m(\\d)c(\\d)", "\\1.\\2", Sample)) %>%
  select(mouse, microbiome.sample, starts_with("norm_"))

colnames(normalized_cleaned_data) <- gsub("^norm_", "", colnames(normalized_cleaned_data))

long_data <- normalized_cleaned_data %>%
  pivot_longer(cols = c(`butyric acid`, `valeric acid`, `heptanoic acid`),
               names_to = "variable", values_to = "value") %>%
  filter(microbiome.sample != "PBS") %>%
  mutate(microbiome.sample = factor(
           microbiome.sample,
           levels = c("Agathobacter rectalis", "Blautia obeum", "Blautia massiliensis")),
         microbiome.sample = recode(microbiome.sample,
                                    "Agathobacter rectalis" = "A. rectalis",
                                    "Blautia obeum" = "B. obeum",
                                    "Blautia massiliensis" = "B. massiliensis"))

ggplot(long_data, aes(x = microbiome.sample, y = log(value), fill = microbiome.sample)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.6) +
  geom_jitter(width = 0.2, size = 1, alpha = 0.8, color = "black") +
  facet_wrap(~variable, scales = "free_y", ncol = 5) +
  labs(x = "", y = "Normalized SCFA") +
  theme_minimal() +
  theme(legend.position = "none",
        strip.text = element_text(size = 10),
        axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_fill_brewer(palette = "Paired", direction = -1)

ggsave("../figures/panels/fig5C_scfa-butyrate-heptanoate-valerate.svg", width = 5, height = 3)
```
