---
title: "16S engraftment analysis"
author: "LN"
date: "2025-11-24"
output: html_document
---

```{r}
getwd()
```


```{r}
install.packages("here")
install.packages("svglite")
install.packages("systemfonts")
```

```{r}
library(tidyverse)
library(vegan)
library(dplyr)
library(here) 

```

```{r}
bewell.relabun <- read.csv("../data/BEWELL/RelAbun_gen.csv")
seqtab.mimic <- readRDS("../data/mimic/2023-07-26_16S-seqtab.rds") %>%
  as.data.frame()
taxa.mimic <- readRDS("../data/mimic/2023-07-26_16S-taxa.rds") %>%
  as.data.frame()
key <- read_csv("../data/mimic/mimic16S-key.csv") %>%
  select(-`...1`)
group.spread1 <- read_csv("../data/mimic/group_spread.csv")
```


#Helper function: compute PCoA scores and merge with sample metadata 
```{r}
pcoa_df <- function(bind_table, key_sub, bewell_id, bewell_timepoint_label = "Bewell") {

  BC <- vegdist(bind_table, method = "bray") %>%
    as.matrix()

  pcoa <- cmdscale(BC, k = 2, eig = TRUE)

  scores <- as.data.frame(pcoa$points) %>%
    rownames_to_column("sampleID") %>%
    rename(PC1 = V1,
           PC2 = V2)

  meta.mimic <- key_sub %>%
    mutate(sampleID = as.character(sampleID),
           origin   = "Mouse")

  meta.bewell <- tibble(
    sampleID = bewell_id,
    timepoint = bewell_timepoint_label,
    gavage = bewell_timepoint_label,
    origin = "Human"
  )

  meta <- bind_rows(meta.mimic, meta.bewell)

  out <- left_join(scores, meta, by = "sampleID")

  # ---- add % variance explained as attributes ----
  eig <- pcoa$eig
  eig_pos <- eig[eig > 0]
  pct <- eig_pos / sum(eig_pos) * 100

  attr(out, "pc1_label") <- paste0("PC1 (", round(pct[1], 1), "%)")
  attr(out, "pc2_label") <- paste0("PC2 (", round(pct[2], 1), "%)")

  out
}
```

#Bind table: mim12 (HONC85-rB)
```{r}
bind.table.85v1 <- bind_rows(bewell.85v1,mim12v1.relabun) %>%
  mutate(across(.cols = -sampleID, .fns = ~coalesce(., 0))) %>%
  column_to_rownames("sampleID") %>%
  as.matrix()
```

#Generate PCoA coordinates for Mimic 12 (85-rB) with standardized timepoint labels
```{r}
# metadata
key.85v1 <- key %>%
  filter(experiment == "mimic 12", gavage == "Pre-BRB") %>%
  select(sampleID, timepoint, gavage)

time_levels <- c(
  "Study Participant",
  "Pre-antibiotics",
  "Pre-gavage",
  "First Anti-PD1/IgG Inj.",
  "End of Study"
)

# make PCoA df (PC1, PC2 + metadata)
pcoa.85v1_raw <- pcoa_df(
  bind_table = bind.table.85v1,
  key_sub    = key.85v1,
  bewell_id  = "85-rB",
  bewell_timepoint_label = "Bewell"
)

# grab REAL variance labels BEFORE mutate drops attributes
pc1_lab <- attr(pcoa.85v1_raw, "pc1_label")
pc2_lab <- attr(pcoa.85v1_raw, "pc2_label")

# recode + factor levels
pcoa.85v1 <- pcoa.85v1_raw %>%
  mutate(
    timepoint = recode(
      timepoint,
      "Bewell"       = "Study Participant",
      "baseline"     = "Pre-antibiotics",
      "Baseline"     = "Pre-antibiotics",
      "gavage"       = "Pre-gavage",
      "Gavage"       = "Pre-gavage",
      "Anti-PD1/IgG" = "First Anti-PD1/IgG Inj.",
      "necropsy"     = "End of Study",
      "Necropsy"     = "End of Study"
    ),
    timepoint = factor(timepoint, levels = time_levels)
  )

```

#Plot mimic 12(85-rB) 
```{r}
muted_cols <- c(
  "Study Participant"         = "#6FB5A5",
  "Pre-antibiotics"           = "#9A9F6B",
  "Pre-gavage"                = "#9CC4E4",
  "First Anti-PD1/IgG Inj."   = "#D9908A",
  "End of Study"              = "#C8A0D6"
)

# Centroids (mouse only)
centroids <- pcoa.85v1 %>%
  filter(origin == "Mouse") %>%
  group_by(timepoint) %>%
  summarize(PC1 = mean(PC1), PC2 = mean(PC2), .groups = "drop")

ggplot(pcoa.85v1, aes(PC1, PC2)) +

  stat_ellipse(
    data = pcoa.85v1 %>% filter(origin == "Mouse"),
    aes(color = timepoint),
    type = "norm",
    level = 0.95,
    alpha = 0.15,
    size = 1,
    show.legend = FALSE
  ) +

  geom_point(
    data = centroids,
    aes(color = timepoint),
    shape = 4,
    size = 5,
    stroke = 1.5,
    show.legend = FALSE
  ) +

 geom_point(aes(color = timepoint, shape = origin, size = origin)) +
scale_size_manual(values = c(Human = 4, Mouse = 3), guide = "none") +

  theme_classic(base_size = 8) +
  coord_equal() +
 labs(
    title = "Distance from human donor (85) to recipient mice (rB) ",
    x = pc1_lab,
    y = pc2_lab,
    color = "Timepoint",
    shape = "Origin"
  ) + 
  theme(plot.title = element_text(hjust = 0.5))


ggsave("../figures/supplement-pcoa-85rB.svg", width = 7, height = 5)
```

#Bind table: mim12 (HONC85-tB)
```{r}
bind.table.85v2 <- bind_rows(bewell.85v2,mim12v2.relabun) %>%
  mutate(across(.cols = -sampleID, .fns = ~coalesce(., 0))) %>%
  column_to_rownames("sampleID") %>%
  as.matrix()
```

#Generate PCoA coordinates for Mimic 12 (85-tB) with standardized timepoint labels
```{r}
# metadata
key.85v2 <- key %>%
  filter(experiment == "mimic 12", gavage == "Post-BRB") %>%
  select(sampleID, timepoint, gavage)

time_levels <- c(
  "Study Participant",
  "Pre-antibiotics",
  "Pre-gavage",
  "First Anti-PD1/IgG Inj.",
  "End of Study"
)

# make PCoA df (PC1, PC2 + metadata)
pcoa.85v2_raw <- pcoa_df(
  bind_table = bind.table.85v2,
  key_sub    = key.85v2,
  bewell_id  = "85-tB",
  bewell_timepoint_label = "Bewell"
)

# grab REAL variance labels BEFORE mutate drops attributes
pc1_lab_85v2 <- attr(pcoa.85v2_raw, "pc1_label")
pc2_lab_85v2 <- attr(pcoa.85v2_raw, "pc2_label")

# recode + factor levels
pcoa.85v2 <- pcoa.85v2_raw %>%
  mutate(
    timepoint = recode(
      timepoint,
      "Bewell"       = "Study Participant",
      "baseline"     = "Pre-antibiotics",
      "Baseline"     = "Pre-antibiotics",
      "gavage"       = "Pre-gavage",
      "Gavage"       = "Pre-gavage",
      "Anti-PD1/IgG" = "First Anti-PD1/IgG Inj.",
      "necropsy"     = "End of Study",
      "Necropsy"     = "End of Study"
    ),
    timepoint = factor(timepoint, levels = time_levels)
  )
```

#Plot mimic 12(85-tB) 
```{r}
muted_cols <- c(
  "Study Participant"         = "#6FB5A5",
  "Pre-antibiotics"           = "#9A9F6B",
  "Pre-gavage"                = "#9CC4E4",
  "First Anti-PD1/IgG Inj."   = "#D9908A",
  "End of Study"              = "#C8A0D6"
)

# Centroids (mouse only)
centroids <- pcoa.85v2 %>%
  filter(origin == "Mouse") %>%
  group_by(timepoint) %>%
  summarize(PC1 = mean(PC1), PC2 = mean(PC2), .groups = "drop")

ggplot(pcoa.85v2, aes(PC1, PC2)) +

  stat_ellipse(
    data = pcoa.85v2 %>% filter(origin == "Mouse"),
    aes(color = timepoint),
    type = "norm",
    level = 0.95,
    alpha = 0.15,
    size = 1,
    show.legend = FALSE
  ) +

  geom_point(
    data = centroids,
    aes(color = timepoint),
    shape = 4,
    size = 5,
    stroke = 1.5,
    show.legend = FALSE
  ) +

  geom_point(aes(color = timepoint, shape = origin, size = origin)) +
  scale_size_manual(values = c(Human = 4, Mouse = 3), guide = "none") +

  theme_classic(base_size = 8) +
  coord_equal() +
  labs(
    title = "Distance from human donor (85) to recipient mice (tB)",
    x = pc1_lab_85v2,
    y = pc2_lab_85v2,
    color = "Timepoint",
    shape = "Origin"
  )

ggsave("../figures/supplement-pcoa-85tB.svg", width = 7, height = 5)
```

#Bind table: mim13 (HONC84-rB)
```{r}
bind.table.84v1 <- bind_rows(bewell.84v1,mim13v1.relabun) %>%
  mutate(across(.cols = -sampleID, .fns = ~coalesce(., 0))) %>%
  column_to_rownames("sampleID") %>%
  as.matrix()
```

#Generate PCoA coordinates for Mimic 13 (84-rB) with standardized timepoint labels
```{r}
# metadata
key.84v1 <- key %>%
  filter(experiment == "mimic 13", gavage == "Pre-BRB") %>%
  select(sampleID, timepoint, gavage)

time_levels <- c(
  "Study Participant",
  "Pre-antibiotics",
  "Pre-gavage",
  "First Anti-PD1/IgG Inj.",
  "End of Study"
)

# make PCoA df (PC1, PC2 + metadata)
pcoa.84v1_raw <- pcoa_df(
  bind_table = bind.table.84v1,
  key_sub    = key.84v1,
  bewell_id  = "84-rB",
  bewell_timepoint_label = "Bewell"
)

# grab REAL variance labels BEFORE mutate drops attributes
pc1_lab_84v1 <- attr(pcoa.84v1_raw, "pc1_label")
pc2_lab_84v1 <- attr(pcoa.84v1_raw, "pc2_label")

# recode + factor levels
pcoa.84v1 <- pcoa.84v1_raw %>%
  mutate(
    timepoint = recode(
      timepoint,
      "Bewell"       = "Study Participant",
      "baseline"     = "Pre-antibiotics",
      "Baseline"     = "Pre-antibiotics",
      "gavage"       = "Pre-gavage",
      "Gavage"       = "Pre-gavage",
      "Anti-PD1/IgG" = "First Anti-PD1/IgG Inj.",
      "necropsy"     = "End of Study",
      "Necropsy"     = "End of Study"
    ),
    timepoint = factor(timepoint, levels = time_levels)
  )
```

#Plot mimic 13(84-rB) 
```{r}
muted_cols <- c(
  "Study Participant"         = "#6FB5A5",
  "Pre-antibiotics"           = "#9A9F6B",
  "Pre-gavage"                = "#9CC4E4",
  "First Anti-PD1/IgG Inj."   = "#D9908A",
  "End of Study"              = "#C8A0D6"
)

# Centroids (mouse only)
centroids <- pcoa.84v1 %>%
  filter(origin == "Mouse") %>%
  group_by(timepoint) %>%
  summarize(PC1 = mean(PC1), PC2 = mean(PC2), .groups = "drop")

ggplot(pcoa.84v1, aes(PC1, PC2)) +

  stat_ellipse(
    data = pcoa.84v1 %>% filter(origin == "Mouse"),
    aes(color = timepoint),
    type = "norm",
    level = 0.95,
    alpha = 0.15,
    size = 1,
    show.legend = FALSE
  ) +

  geom_point(
    data = centroids,
    aes(color = timepoint),
    shape = 4,
    size = 5,
    stroke = 1.5,
    show.legend = FALSE
  ) +

  geom_point(aes(color = timepoint, shape = origin, size = origin)) +
  scale_size_manual(values = c(Human = 4, Mouse = 3), guide = "none") +

  theme_classic(base_size = 8) +
  coord_equal() +
  labs(
    title = "Distance from human donor (84) to recipient mice (rB)",
    x = pc1_lab_84v1,
    y = pc2_lab_84v1,
    color = "Timepoint",
    shape = "Origin"
  ) + 
  theme(plot.title = element_text(hjust = 0.5))

ggsave("../figures/supplement-pcoa-84rB.png", width = 7, height = 5)
```

#Bind table: mim13 (HONC84-tB)
```{r}
bind.table.84v2 <- bind_rows(bewell.84v2,mim13v2.relabun) %>%
  mutate(across(.cols = -sampleID, .fns = ~coalesce(., 0))) %>%
  column_to_rownames("sampleID") %>%
  as.matrix()
```

#Generate PCoA coordinates for Mimic 13 (84-tB) with standardized timepoint labels
```{r}
# metadata
key.84v2 <- key %>%
  filter(experiment == "mimic 13", gavage == "Post-BRB") %>%
  select(sampleID, timepoint, gavage)

time_levels <- c(
  "Study Participant",
  "Pre-antibiotics",
  "Pre-gavage",
  "First Anti-PD1/IgG Inj.",
  "End of Study"
)

# make PCoA df (PC1, PC2 + metadata)
pcoa.84v2_raw <- pcoa_df(
  bind_table = bind.table.84v2,
  key_sub    = key.84v2,
  bewell_id  = "84-tB",
  bewell_timepoint_label = "Bewell"
)

# grab REAL variance labels BEFORE mutate drops attributes
pc1_lab_84v2 <- attr(pcoa.84v2_raw, "pc1_label")
pc2_lab_84v2 <- attr(pcoa.84v2_raw, "pc2_label")

# recode + factor levels
pcoa.84v2 <- pcoa.84v2_raw %>%
  mutate(
    timepoint = recode(
      timepoint,
      "Bewell"       = "Study Participant",
      "baseline"     = "Pre-antibiotics",
      "Baseline"     = "Pre-antibiotics",
      "gavage"       = "Pre-gavage",
      "Gavage"       = "Pre-gavage",
      "Anti-PD1/IgG" = "First Anti-PD1/IgG Inj.",
      "necropsy"     = "End of Study",
      "Necropsy"     = "End of Study"
    ),
    timepoint = factor(timepoint, levels = time_levels)
  )
```

#Plot mimic 13(84-tB) 
```{r}
muted_cols <- c(
  "Study Participant"         = "#6FB5A5",
  "Pre-antibiotics"           = "#9A9F6B",
  "Pre-gavage"                = "#9CC4E4",
  "First Anti-PD1/IgG Inj."   = "#D9908A",
  "End of Study"              = "#C8A0D6"
)

# Centroids (mouse only)
centroids <- pcoa.84v2 %>%
  filter(origin == "Mouse") %>%
  group_by(timepoint) %>%
  summarize(PC1 = mean(PC1), PC2 = mean(PC2), .groups = "drop")

ggplot(pcoa.84v2, aes(PC1, PC2)) +

  stat_ellipse(
    data = pcoa.84v2 %>% filter(origin == "Mouse"),
    aes(color = timepoint),
    type = "norm",
    level = 0.95,
    alpha = 0.15,
    size = 1,
    show.legend = FALSE
  ) +

  geom_point(
    data = centroids,
    aes(color = timepoint),
    shape = 4,
    size = 5,
    stroke = 1.5,
    show.legend = FALSE
  ) +

  geom_point(aes(color = timepoint, shape = origin, size = origin)) +
  scale_size_manual(values = c(Human = 4, Mouse = 3), guide = "none") +

  theme_classic(base_size = 8) +
  coord_equal() +
  labs(
    title = "Distance from human donor (84) to recipient mice (tB)",
    x = pc1_lab_84v2,
    y = pc2_lab_84v2,
    color = "Timepoint",
    shape = "Origin"
  )+ 
  theme(plot.title = element_text(hjust = 0.5))

ggsave("../figures/supplement-pcoa-84tB.svg", width = 7, height = 5)
```

#Bind table: mim10 (HONC68-rB)
```{r}
bind.table.68v1 <- bind_rows(bewell.68v1,mim10v1.relabun) %>%
  mutate(across(.cols = -sampleID, .fns = ~coalesce(., 0))) %>%
  column_to_rownames("sampleID") %>%
  as.matrix()
```

#Generate PCoA coordinates for Mimic 10 (68-rB) with standardized timepoint labels
```{r}
# metadata
key.68v1 <- key %>%
  filter(experiment == "mimic 10", gavage == "Pre-BRB") %>%
  select(sampleID, timepoint, gavage)

time_levels <- c(
  "Study Participant",
  "Pre-antibiotics",
  "Pre-gavage",
  "First Anti-PD1/IgG Inj.",
  "End of Study"
)

# START FRESH: run pcoa_df once, capture variance labels, then recode
pcoa.68v1_raw <- pcoa_df(
  bind_table = bind.table.68v1,
  key_sub    = key.68v1,
  bewell_id  = "68-rB",
  bewell_timepoint_label = "Bewell"
)

# grab REAL variance labels BEFORE mutate drops attributes
pc1_lab_68v1 <- attr(pcoa.68v1_raw, "pc1_label")
pc2_lab_68v1 <- attr(pcoa.68v1_raw, "pc2_label")

pcoa.68v1 <- pcoa.68v1_raw %>%
  mutate(
    # preserve ORIGINAL strings before recoding
    timepoint_raw = trimws(tolower(as.character(timepoint))),
    gavage_raw    = trimws(tolower(as.character(gavage))),

    timepoint = case_when(
      origin == "Human" ~ "Study Participant",
      timepoint_raw == "baseline" ~ "Pre-antibiotics",
      timepoint_raw %in% c("anti-pd1/igg", "antipd1/igg") ~ "First Anti-PD1/IgG Inj.",
      timepoint_raw == "necropsy" ~ "End of Study",
      timepoint_raw %in% c("gavage", "mc38") ~ "Pre-gavage",
      # ONLY use gavage as fallback when timepoint is missing/blank
      (is.na(timepoint_raw) | timepoint_raw == "") & gavage_raw %in% c("pre-brb", "post-brb") ~ "Pre-gavage",
      TRUE ~ NA_character_
    ),
    timepoint = factor(timepoint, levels = time_levels)
  )

```

#Plot mimic 10(68-rB) 
```{r}
muted_cols <- c(
  "Study Participant"         = "#6FB5A5",
  "Pre-antibiotics"           = "#9A9F6B",
  "Pre-gavage"                = "#9CC4E4",
  "First Anti-PD1/IgG Inj."   = "#D9908A",
  "End of Study"              = "#C8A0D6"
)

# Centroids (mouse only)
centroids <- pcoa.68v1 %>%
  filter(origin == "Mouse") %>%
  group_by(timepoint) %>%
  summarize(PC1 = mean(PC1), PC2 = mean(PC2), .groups = "drop")

ggplot(pcoa.68v1, aes(PC1, PC2)) +

  stat_ellipse(
    data = pcoa.68v1 %>% filter(origin == "Mouse"),
    aes(color = timepoint),
    type = "norm",
    level = 0.95,
    alpha = 0.15,
    size = 1,
    show.legend = FALSE
  ) +

  geom_point(
    data = centroids,
    aes(color = timepoint),
    shape = 4,
    size = 5,
    stroke = 1.5,
    show.legend = FALSE
  ) +

  geom_point(aes(color = timepoint, shape = origin, size = origin)) +
  scale_size_manual(values = c(Human = 4, Mouse = 3), guide = "none") +

  theme_classic(base_size = 8) +
  coord_equal() +
  labs(
    title = "Distance from human donor (68) to recipient mice (rB)",
    x = pc1_lab_68v1,
    y = pc2_lab_68v1,
    color = "Timepoint",
    shape = "Origin"
  )+ 
  theme(plot.title = element_text(hjust = 0.5))

ggsave("../figures/supplement-pcoa-68rB.svg", width = 7, height = 5)
```


#Bind table: mim10 (HONC68-tB)
```{r}
bind.table.68v2 <- bind_rows(bewell.68v2,mim10v2.relabun) %>%
  mutate(across(.cols = -sampleID, .fns = ~coalesce(., 0))) %>%
  column_to_rownames("sampleID") %>%
  as.matrix()
```

#Generate PCoA coordinates for Mimic 10 (68-tB) with standardized timepoint labels
```{r}
# metadata
key.68v2 <- key %>%
  filter(experiment == "mimic 10", gavage == "Post-BRB") %>%
  select(sampleID, timepoint, gavage)

time_levels <- c(
  "Study Participant",
  "Pre-antibiotics",
  "Pre-gavage",
  "First Anti-PD1/IgG Inj.",
  "End of Study"
)

# START FRESH: run pcoa_df once, capture variance labels, then recode
pcoa.68v2_raw <- pcoa_df(
  bind_table = bind.table.68v2,
  key_sub    = key.68v2,
  bewell_id  = "68-tB",
  bewell_timepoint_label = "Bewell"
)

# grab REAL variance labels BEFORE mutate drops attributes
pc1_lab_68v2 <- attr(pcoa.68v2_raw, "pc1_label")
pc2_lab_68v2 <- attr(pcoa.68v2_raw, "pc2_label")

pcoa.68v2 <- pcoa.68v2_raw %>%
  mutate(
    # preserve ORIGINAL strings before recoding
    timepoint_raw = trimws(tolower(as.character(timepoint))),
    gavage_raw    = trimws(tolower(as.character(gavage))),

    timepoint = case_when(
      origin == "Human" ~ "Study Participant",
      timepoint_raw == "baseline" ~ "Pre-antibiotics",
      timepoint_raw %in% c("anti-pd1/igg", "antipd1/igg") ~ "First Anti-PD1/IgG Inj.",
      timepoint_raw == "necropsy" ~ "End of Study",
      timepoint_raw %in% c("gavage", "mc38") ~ "Pre-gavage",
      # ONLY use gavage as fallback when timepoint is missing/blank
      (is.na(timepoint_raw) | timepoint_raw == "") & gavage_raw %in% c("pre-brb", "post-brb") ~ "Pre-gavage",
      TRUE ~ NA_character_
    ),
    timepoint = factor(timepoint, levels = time_levels)
  )

```

#Plot mimic 10(68-tB) 
```{r}
muted_cols <- c(
  "Study Participant"         = "#6FB5A5",
  "Pre-antibiotics"           = "#9A9F6B",
  "Pre-gavage"                = "#9CC4E4",
  "First Anti-PD1/IgG Inj."   = "#D9908A",
  "End of Study"              = "#C8A0D6"
)

# Centroids (mouse only)
centroids <- pcoa.68v2 %>%
  filter(origin == "Mouse") %>%
  group_by(timepoint) %>%
  summarize(PC1 = mean(PC1), PC2 = mean(PC2), .groups = "drop")

ggplot(pcoa.68v2, aes(PC1, PC2)) +

  stat_ellipse(
    data = pcoa.68v2 %>% filter(origin == "Mouse"),
    aes(color = timepoint),
    type = "norm",
    level = 0.95,
    alpha = 0.15,
    size = 1,
    show.legend = FALSE
  ) +

  geom_point(
    data = centroids,
    aes(color = timepoint),
    shape = 4,
    size = 5,
    stroke = 1.5,
    show.legend = FALSE
  ) +

  geom_point(aes(color = timepoint, shape = origin, size = origin)) +
  scale_size_manual(values = c(Human = 4, Mouse = 3), guide = "none") +

  theme_classic(base_size = 8) +
  coord_equal() +
  labs(
    title = "Distance from human donor (68) to recipient mice (tB)",
    x = pc1_lab_68v2,
    y = pc2_lab_68v2,
    color = "Timepoint",
    shape = "Origin"
  )+ 
  theme(plot.title = element_text(hjust = 0.5))

ggsave("../figures/supplement-pcoa-68tB.svg", width = 7, height = 5)
```

#Bind table: mim11 (HONC79-tB)
```{r}
bind.table.79v2 <- bind_rows(bewell.79v2,mim11v2.relabun) %>%
  mutate(across(.cols = -sampleID, .fns = ~coalesce(., 0))) %>%
  column_to_rownames("sampleID") %>%
  as.matrix()
```

#Generate PCoA coordinates for Mimic 11 (79-tB) with standardized timepoint labels
```{r}
# metadata
key.79v2 <- key %>%
  filter(experiment == "mimic 11", gavage == "Post-BRB") %>%
  select(sampleID, timepoint, gavage)

time_levels <- c(
  "Study Participant",
  "Pre-antibiotics",
  "Pre-gavage",
  "First Anti-PD1/IgG Inj.",
  "End of Study"
)

# START FRESH: run pcoa_df once, capture variance labels, then recode
pcoa.79v2_raw <- pcoa_df(
  bind_table = bind.table.79v2,
  key_sub    = key.79v2,
  bewell_id  = "79-tB",
  bewell_timepoint_label = "Bewell"
)

# grab REAL variance labels BEFORE mutate drops attributes
pc1_lab_79v2 <- attr(pcoa.79v2_raw, "pc1_label")
pc2_lab_79v2 <- attr(pcoa.79v2_raw, "pc2_label")

pcoa.79v2 <- pcoa.79v2_raw %>%
  mutate(
    # preserve ORIGINAL strings before recoding
    timepoint_raw = trimws(tolower(as.character(timepoint))),
    gavage_raw    = trimws(tolower(as.character(gavage))),

    timepoint = case_when(
      origin == "Human" ~ "Study Participant",
      timepoint_raw == "baseline" ~ "Pre-antibiotics",
      timepoint_raw %in% c("anti-pd1/igg", "antipd1/igg") ~ "First Anti-PD1/IgG Inj.",
      timepoint_raw == "necropsy" ~ "End of Study",
      timepoint_raw %in% c("gavage", "mc38") ~ "Pre-gavage",
      # ONLY use gavage as fallback when timepoint is missing/blank
      (is.na(timepoint_raw) | timepoint_raw == "") & gavage_raw %in% c("pre-brb", "post-brb") ~ "Pre-gavage",
      TRUE ~ NA_character_
    ),
    timepoint = factor(timepoint, levels = time_levels)
  )

```

#Plot mimic 11(79-tB) 
```{r}
muted_cols <- c(
  "Study Participant"         = "#6FB5A5",
  "Pre-antibiotics"           = "#9A9F6B",
  "Pre-gavage"                = "#9CC4E4",
  "First Anti-PD1/IgG Inj."   = "#D9908A",
  "End of Study"              = "#C8A0D6"
)

# Centroids (mouse only)
centroids <- pcoa.79v2 %>%
  filter(origin == "Mouse") %>%
  group_by(timepoint) %>%
  summarize(PC1 = mean(PC1), PC2 = mean(PC2), .groups = "drop")

ggplot(pcoa.79v2, aes(PC1, PC2)) +

  stat_ellipse(
    data = pcoa.79v2 %>% filter(origin == "Mouse"),
    aes(color = timepoint),
    type = "norm",
    level = 0.95,
    alpha = 0.15,
    size = 1,
    show.legend = FALSE
  ) +

  geom_point(
    data = centroids,
    aes(color = timepoint),
    shape = 4,
    size = 5,
    stroke = 1.5,
    show.legend = FALSE
  ) +

  geom_point(aes(color = timepoint, shape = origin, size = origin)) +
  scale_size_manual(values = c(Human = 4, Mouse = 3), guide = "none") +

  theme_classic(base_size = 8) +
  coord_equal() +
  labs(
    title = "Distance from human donor (79) to recipient mice (tB)",
    x = pc1_lab_79v2,
    y = pc2_lab_79v2,
    color = "Timepoint",
    shape = "Origin"
  )+ 
  theme(plot.title = element_text(hjust = 0.5))

ggsave("../figures/supplement-pcoa-79tB.svg", width = 7, height = 5)
```
#Bind table: mim5 (HONC102-rB)
```{r}
bind.table.102v1 <- bind_rows(bewell.102v1,mim5v1.relabun) %>%
  mutate(across(.cols = -sampleID, .fns = ~coalesce(., 0))) %>%
  column_to_rownames("sampleID") %>%
  as.matrix()
```

#Generate PCoA coordinates for Mimic 5 (102-rB) with standardized timepoint labels
```{r}
# metadata
key.102v1 <- key %>%
  filter(experiment == "mimic 5", gavage == "Pre-BRB") %>%
  select(sampleID, timepoint, gavage)

time_levels <- c(
  "Study Participant",
  "Pre-antibiotics",
  "Pre-gavage",
  "First Anti-PD1/IgG Inj.",
  "End of Study"
)

# START FRESH: run pcoa_df once, capture variance labels, then recode
pcoa.102v1_raw <- pcoa_df(
  bind_table = bind.table.102v1,
  key_sub    = key.102v1,
  bewell_id  = "102-rB",
  bewell_timepoint_label = "Bewell"
)

# grab REAL variance labels BEFORE mutate drops attributes
pc1_lab_102v1 <- attr(pcoa.102v1_raw, "pc1_label")
pc2_lab_102v1 <- attr(pcoa.102v1_raw, "pc2_label")

pcoa.102v1 <- pcoa.102v1_raw %>%
  mutate(
    # preserve ORIGINAL strings before recoding
    timepoint_raw = trimws(tolower(as.character(timepoint))),
    gavage_raw    = trimws(tolower(as.character(gavage))),

    timepoint = case_when(
      origin == "Human" ~ "Study Participant",
      timepoint_raw == "baseline" ~ "Pre-antibiotics",
      timepoint_raw %in% c("anti-pd1/igg", "antipd1/igg") ~ "First Anti-PD1/IgG Inj.",
      timepoint_raw == "necropsy" ~ "End of Study",
      timepoint_raw %in% c("gavage", "mc38") ~ "Pre-gavage",
      # ONLY use gavage as fallback when timepoint is missing/blank
      (is.na(timepoint_raw) | timepoint_raw == "") & gavage_raw %in% c("pre-brb", "post-brb") ~ "Pre-gavage",
      TRUE ~ NA_character_
    ),
    timepoint = factor(timepoint, levels = time_levels)
  )

```

#Plot mimic 5(102-rB) 
```{r}
muted_cols <- c(
  "Study Participant"         = "#6FB5A5",
  "Pre-antibiotics"           = "#9A9F6B",
  "Pre-gavage"                = "#9CC4E4",
  "First Anti-PD1/IgG Inj."   = "#D9908A",
  "End of Study"              = "#C8A0D6"
)

# Centroids (mouse only)
centroids <- pcoa.102v1 %>%
  filter(origin == "Mouse") %>%
  group_by(timepoint) %>%
  summarize(PC1 = mean(PC1), PC2 = mean(PC2), .groups = "drop")

ggplot(pcoa.102v1, aes(PC1, PC2)) +

  stat_ellipse(
    data = pcoa.102v1 %>% filter(origin == "Mouse"),
    aes(color = timepoint),
    type = "norm",
    level = 0.95,
    alpha = 0.15,
    size = 1,
    show.legend = FALSE
  ) +

  geom_point(
    data = centroids,
    aes(color = timepoint),
    shape = 4,
    size = 5,
    stroke = 1.5,
    show.legend = FALSE
  ) +

  geom_point(aes(color = timepoint, shape = origin, size = origin)) +
  scale_size_manual(values = c(Human = 4, Mouse = 3), guide = "none") +

  theme_classic(base_size = 8) +
  coord_equal() +
  labs(
    title = "Distance from human donor (102) to recipient mice (rB)",
    x = pc1_lab_102v1,
    y = pc2_lab_102v1,
    color = "Timepoint",
    shape = "Origin"
  ) +
  theme(plot.title = element_text(hjust = 0.5))

ggsave("../figures/supplement-pcoa-102rB.svg", width = 7, height = 5)
```

#Bind table: mim5 (HONC102-tB)
```{r}
bind.table.102v2 <- bind_rows(bewell.102v2,mim5v2.relabun) %>%
  mutate(across(.cols = -sampleID, .fns = ~coalesce(., 0))) %>%
  column_to_rownames("sampleID") %>%
  as.matrix()
```

#Generate PCoA coordinates for Mimic 5 (102-rB) with standardized timepoint labels
```{r}
# metadata
key.102v2 <- key %>%
  filter(experiment == "mimic 5", gavage == "Post-BRB") %>%
  select(sampleID, timepoint, gavage)

time_levels <- c(
  "Study Participant",
  "Pre-antibiotics",
  "Pre-gavage",
  "First Anti-PD1/IgG Inj.",
  "End of Study"
)

# START FRESH: run pcoa_df once, capture variance labels, then recode
pcoa.102v2_raw <- pcoa_df(
  bind_table = bind.table.102v2,
  key_sub    = key.102v2,
  bewell_id  = "102-tB",
  bewell_timepoint_label = "Bewell"
)

# grab REAL variance labels BEFORE mutate drops attributes
pc1_lab_102v2 <- attr(pcoa.102v2_raw, "pc1_label")
pc2_lab_102v2 <- attr(pcoa.102v2_raw, "pc2_label")

pcoa.102v2 <- pcoa.102v2_raw %>%
  mutate(
    # preserve ORIGINAL strings before recoding
    timepoint_raw = trimws(tolower(as.character(timepoint))),
    gavage_raw    = trimws(tolower(as.character(gavage))),

    timepoint = case_when(
      origin == "Human" ~ "Study Participant",
      timepoint_raw == "baseline" ~ "Pre-antibiotics",
      timepoint_raw %in% c("anti-pd1/igg", "antipd1/igg") ~ "First Anti-PD1/IgG Inj.",
      timepoint_raw == "necropsy" ~ "End of Study",
      timepoint_raw %in% c("gavage", "mc38") ~ "Pre-gavage",
      # ONLY use gavage as fallback when timepoint is missing/blank
      (is.na(timepoint_raw) | timepoint_raw == "") & gavage_raw %in% c("pre-brb", "post-brb") ~ "Pre-gavage",
      TRUE ~ NA_character_
    ),
    timepoint = factor(timepoint, levels = time_levels)
  )

```

#Plot mimic 5(102-tB) 
```{r}
muted_cols <- c(
  "Study Participant"         = "#6FB5A5",
  "Pre-antibiotics"           = "#9A9F6B",
  "Pre-gavage"                = "#9CC4E4",
  "First Anti-PD1/IgG Inj."   = "#D9908A",
  "End of Study"              = "#C8A0D6"
)

# Centroids (mouse only)
centroids <- pcoa.102v2 %>%
  filter(origin == "Mouse") %>%
  group_by(timepoint) %>%
  summarize(PC1 = mean(PC1), PC2 = mean(PC2), .groups = "drop")

ggplot(pcoa.102v2, aes(PC1, PC2)) +

  stat_ellipse(
    data = pcoa.102v2 %>% filter(origin == "Mouse"),
    aes(color = timepoint),
    type = "norm",
    level = 0.95,
    alpha = 0.15,
    size = 1,
    show.legend = FALSE
  ) +

  geom_point(
    data = centroids,
    aes(color = timepoint),
    shape = 4,
    size = 5,
    stroke = 1.5,
    show.legend = FALSE
  ) +

  geom_point(aes(color = timepoint, shape = origin, size = origin)) +
  scale_size_manual(values = c(Human = 4, Mouse = 3), guide = "none") +
  scale_color_manual(values = muted_cols, drop = TRUE) +

  theme_classic(base_size = 8) +
  coord_equal() +
  labs(
    title = "Distance from human donor (102) to recipient mice (tB)",
    x = pc1_lab_102v2,
    y = pc2_lab_102v2,
    color = "Timepoint",
    shape = "Origin"
  )+theme(plot.title = element_text(hjust = 0.5))
  

ggsave("../figures/supplement-pcoa-102tB.svg", width = 7, height = 5)
```

#Bind table: mim1 (HONC55-rB)
```{r}
bind.table.55v1 <- bind_rows(bewell.55v1,mim1v1.relabun) %>%
  mutate(across(.cols = -sampleID, .fns = ~coalesce(., 0))) %>%
  column_to_rownames("sampleID") %>%
  as.matrix()
```

#Generate PCoA coordinates for Mimic 1 (55-rB) with standardized timepoint labels
```{r}
# metadata
key.55v1 <- key %>%
  filter(experiment == "mimic 1", gavage == "Pre-BRB") %>%
  select(sampleID, timepoint, gavage)

time_levels <- c(
  "Study Participant",
  "Pre-antibiotics",
  "Pre-gavage",
  "First Anti-PD1/IgG Inj.",
  "End of Study"
)

# START FRESH: run pcoa_df once, capture variance labels, then recode
pcoa.55v1_raw <- pcoa_df(
  bind_table = bind.table.55v1,
  key_sub    = key.55v1,
  bewell_id  = "55-rB",
  bewell_timepoint_label = "Bewell"
)

# grab REAL variance labels BEFORE mutate drops attributes
pc1_lab_55v1 <- attr(pcoa.55v1_raw, "pc1_label")
pc2_lab_55v1 <- attr(pcoa.55v1_raw, "pc2_label")

pcoa.55v1 <- pcoa.55v1_raw %>%
  mutate(
    # preserve ORIGINAL strings before recoding
    timepoint_raw = trimws(tolower(as.character(timepoint))),
    gavage_raw    = trimws(tolower(as.character(gavage))),

    timepoint = case_when(
      origin == "Human" ~ "Study Participant",
      timepoint_raw == "baseline" ~ "Pre-antibiotics",
      timepoint_raw %in% c("anti-pd1/igg", "antipd1/igg") ~ "First Anti-PD1/IgG Inj.",
      timepoint_raw == "necropsy" ~ "End of Study",
      timepoint_raw %in% c("gavage", "mc38") ~ "Pre-gavage",
      # ONLY use gavage as fallback when timepoint is missing/blank
      (is.na(timepoint_raw) | timepoint_raw == "") & gavage_raw %in% c("pre-brb", "post-brb") ~ "Pre-gavage",
      TRUE ~ NA_character_
    ),
    timepoint = factor(timepoint, levels = time_levels)
  )

```

#Plot mimic 1(55-rB) 
```{r}
muted_cols <- c(
  "Study Participant"         = "#6FB5A5",
  "Pre-antibiotics"           = "#9A9F6B",
  "Pre-gavage"                = "#9CC4E4",
  "First Anti-PD1/IgG Inj."   = "#D9908A",
  "End of Study"              = "#C8A0D6"
)

# Centroids (mouse only)
centroids <- pcoa.55v1 %>%
  filter(origin == "Mouse") %>%
  group_by(timepoint) %>%
  summarize(PC1 = mean(PC1), PC2 = mean(PC2), .groups = "drop")

ggplot(pcoa.55v1, aes(PC1, PC2)) +

  stat_ellipse(
    data = pcoa.55v1 %>% filter(origin == "Mouse"),
    aes(color = timepoint),
    type = "norm",
    level = 0.95,
    alpha = 0.15,
    size = 1,
    show.legend = FALSE
  ) +

  geom_point(
    data = centroids,
    aes(color = timepoint),
    shape = 4,
    size = 5,
    stroke = 1.5,
    show.legend = FALSE
  ) +

  geom_point(aes(color = timepoint, shape = origin, size = origin)) +
  scale_size_manual(values = c(Human = 4, Mouse = 3), guide = "none") +
  scale_color_manual(values = muted_cols, drop = TRUE) +

  theme_classic(base_size = 8) +
  coord_equal() +
  labs(
    title = "Distance from human donor (55) to recipient mice (rB)",
    x = pc1_lab_55v1,
    y = pc2_lab_55v1,
    color = "Timepoint",
    shape = "Origin"
  )+
  theme(plot.title = element_text(hjust = 0.5))

ggsave("../figures/supplement-pcoa-55rB.png", width = 7, height = 5)
```

#Bind table: mim1 (HONC55-tB)
```{r}
bind.table.55v2 <- bind_rows(bewell.55v2,mim1v2.relabun) %>%
  mutate(across(.cols = -sampleID, .fns = ~coalesce(., 0))) %>%
  column_to_rownames("sampleID") %>%
  as.matrix()
```

#Generate PCoA coordinates for Mimic 1 (55-tB) with standardized timepoint labels
```{r}
# metadata
key.55v2 <- key %>%
  filter(experiment == "mimic 1", gavage == "Post-BRB") %>%
  select(sampleID, timepoint, gavage)

time_levels <- c(
  "Study Participant",
  "Pre-antibiotics",
  "Pre-gavage",
  "First Anti-PD1/IgG Inj.",
  "End of Study"
)

# START FRESH: run pcoa_df once, capture variance labels, then recode
pcoa.55v2_raw <- pcoa_df(
  bind_table = bind.table.55v2,
  key_sub    = key.55v2,
  bewell_id  = "55-tB",
  bewell_timepoint_label = "Bewell"
)

# grab REAL variance labels BEFORE mutate drops attributes
pc1_lab_55v2 <- attr(pcoa.55v2_raw, "pc1_label")
pc2_lab_55v2 <- attr(pcoa.55v2_raw, "pc2_label")

pcoa.55v2 <- pcoa.55v2_raw %>%
  mutate(
    # preserve ORIGINAL strings before recoding
    timepoint_raw = trimws(tolower(as.character(timepoint))),
    gavage_raw    = trimws(tolower(as.character(gavage))),

    timepoint = case_when(
      origin == "Human" ~ "Study Participant",
      timepoint_raw == "baseline" ~ "Pre-antibiotics",
      timepoint_raw %in% c("anti-pd1/igg", "antipd1/igg") ~ "First Anti-PD1/IgG Inj.",
      timepoint_raw == "necropsy" ~ "End of Study",
      timepoint_raw %in% c("gavage", "mc38") ~ "Pre-gavage",
      # ONLY use gavage as fallback when timepoint is missing/blank
      (is.na(timepoint_raw) | timepoint_raw == "") & gavage_raw %in% c("pre-brb", "post-brb") ~ "Pre-gavage",
      TRUE ~ NA_character_
    ),
    timepoint = factor(timepoint, levels = time_levels)
  )

```

#Plot mimic 1(55-tB) 
```{r}
muted_cols <- c(
  "Study Participant"         = "#6FB5A5",
  "Pre-antibiotics"           = "#9A9F6B",
  "Pre-gavage"                = "#9CC4E4",
  "First Anti-PD1/IgG Inj."   = "#D9908A",
  "End of Study"              = "#C8A0D6"
)

# Centroids (mouse only)
centroids <- pcoa.55v2 %>%
  filter(origin == "Mouse") %>%
  group_by(timepoint) %>%
  summarize(PC1 = mean(PC1), PC2 = mean(PC2), .groups = "drop")

ggplot(pcoa.55v2, aes(PC1, PC2)) +

  stat_ellipse(
    data = pcoa.55v2 %>% filter(origin == "Mouse"),
    aes(color = timepoint),
    type = "norm",
    level = 0.95,
    alpha = 0.15,
    size = 1,
    show.legend = FALSE
  ) +

  geom_point(
    data = centroids,
    aes(color = timepoint),
    shape = 4,
    size = 5,
    stroke = 1.5,
    show.legend = FALSE
  ) +

  geom_point(aes(color = timepoint, shape = origin, size = origin)) +
  scale_size_manual(values = c(Human = 4, Mouse = 3), guide = "none") +
  scale_color_manual(values = muted_cols, drop = TRUE) +

  theme_classic(base_size = 8) +
  coord_equal() +
  labs(
    title = "Distance from human donor (55) to recipient mice (tB)",
    x = pc1_lab_55v2,
    y = pc2_lab_55v2,
    color = "Timepoint",
    shape = "Origin"
  )+theme(plot.title = element_text(hjust = 0.5))

ggsave("../figures/supplement-pcoa-55tB.svg", width = 7, height = 5)
  
```


