---
title: "lachno PCA 16S"
author: "LN"
date: "2025-09-26"
output: html_document
---
```{r}
library(tidyverse)
library(vegan)
```

sample = c("mim16.m2c3t", "mim16.m1c3t", "mim16.m3c3t", "mim16.m1c4t", "mim16.m2c4t", "mim16.m3c4t"),
#Split metadata vs taxa for mimic 11 
```{r}
metadata.11 <- relabun.lachno11 %>%
  select(sampleID, timepoint, gavage.simple, treatment)

otu.11 <- relabun.lachno11 %>%
  select(-timepoint, -gavage.simple, -treatment, -sampleID) %>% 
  as.data.frame()

rownames(otu.11) <- relabun.lachno11$sampleID

```

#Compute distances + PCA for mimic 11  
```{r}
dist_bray <- vegdist(otu.11, method = "bray")

ord.11 <- cmdscale(dist_bray, k=2, eig = TRUE)
ordination_df.11 <- data.frame(
  sampleID = rownames(ord.11$points),
  PC1 = ord.11$points[,1],
  PC2 = ord.11$points[,2]
) %>%
  left_join(metadata.11, by = "sampleID")
```
#compute centroids for mimic 11
```{r}
centroids_11 <- ordination_df.11 %>%
  group_by(timepoint) %>%
  summarize(PC1_centroid = mean(PC1),
            PC2_centroid = mean(PC2))

ordination_df.11 <- ordination_df.11 %>%
  left_join(centroids_11, by = "timepoint") %>%
  mutate(dist_to_centroid = sqrt((PC1 - PC1_centroid)^2 + (PC2 - PC2_centroid)^2))
```

#Plot for mimic 11 
```{r}
ggplot(ordination_df.11, aes(PC1, PC2, color = timepoint)) +
  geom_segment(aes(xend = PC1_centroid, yend = PC2_centroid),
               alpha = 0.4, linewidth = 0.4) +
  geom_point(size = 3, alpha = 0.8) +
  stat_ellipse(type = "t", level = 0.68, linewidth = 0.8) +
  theme_bw() +
  labs(title = "PCoA - Bray–Curtis (mimic 11)",
       x = "PCoA 1", y = "PCoA 2")
```
#plot distances as a separate summary- mimic 11 
```{r}
ggplot(ordination_df.11, aes(x = timepoint, y = dist_to_centroid, fill = timepoint)) +
  geom_boxplot(alpha = 0.6) +
  geom_jitter(width = 0.1, alpha = 0.7) +
  theme_bw() +
  labs(y = "Distance to centroid (Euclidean)",
       x = "Timepoint",
       title = "Group dispersion (mimic 11)")
```


#split metadata vs taxa for mimic 13
```{r}
metadata.13 <- relabun.lachno13 %>%
  select(sampleID, timepoint)

otu <- as.matrix(otu.13)
mode(otu) <- "numeric"
otu[is.na(otu)] <- 0
otu[otu < 0] <- 0

keep_samples <- rowSums(otu) > 0
otu <- otu[keep_samples, , drop = FALSE]
metadata.13 <- metadata.13 %>% filter(sampleID %in% rownames(otu))
otu <- otu[, colSums(otu) > 0, drop = FALSE]
```

#Compute distances + PCA for mimic 13
```{r}
dist_bray_13 <- vegdist(otu, method = "bray")
ord_13 <- cmdscale(dist_bray_13, k = 2, eig = TRUE)

ordination_df_13 <- data.frame(
  sampleID = rownames(ord_13$points),
  PC1 = ord_13$points[,1],
  PC2 = ord_13$points[,2]
) %>% left_join(metadata.13, by = "sampleID")
```

#compute centroids for each group
```{r}
centroids_13 <- ordination_df_13 %>%
  group_by(timepoint) %>%
  summarize(PC1_centroid = mean(PC1),
            PC2_centroid = mean(PC2))

ordination_df_13 <- ordination_df_13 %>%
  left_join(centroids_13, by = "timepoint") %>%
  mutate(dist_to_centroid = sqrt((PC1 - PC1_centroid)^2 + (PC2 - PC2_centroid)^2))
```

#plot PCA for mimic 13
```{r}
ggplot(ordination_df_13, aes(PC1, PC2, color = timepoint)) +
  geom_segment(aes(xend = PC1_centroid, yend = PC2_centroid),
               alpha = 0.4, linewidth = 0.4) +
  geom_point(size = 3, alpha = 0.8) +
  stat_ellipse(type = "t", level = 0.68, linewidth = 0.8) +
  theme_bw() +
  labs(title = "PCoA - Bray–Curtis (mimic 13)",
       x = "PCoA 1", y = "PCoA 2")
```

#plot distances as a separate summary- mimic 13 
```{r}
ggplot(ordination_df_13, aes(x = timepoint, y = dist_to_centroid, fill = timepoint)) +
  geom_boxplot(alpha = 0.6) +
  geom_jitter(width = 0.1, alpha = 0.7) +
  theme_bw() +
  labs(y = "Distance to centroid (Euclidean)",
       x = "Timepoint",
       title = "Group dispersion (mimic 13)")
```



#split metadata vs taxa for mimic 12
```{r}
key_12 <- key_12 %>%
  mutate(
    timepoint = recode(timepoint, 
                       "gavage" = "Post-abx", 
                       "baseline" = "Baseline")
  )

relabun.lachno12 <- key_12 %>%
  left_join(group.spread) %>%
  select(timepoint,gavage.simple,treatment, sampleID, 
         matches("Lachnospiraceae", ignore.case = TRUE))

relabun.lachno12 <- relabun.lachno12 %>%
  mutate(Lachnospiraceae = rowSums(select(., matches("Lachnospiraceae", ignore.case = TRUE)),
                                   na.rm = TRUE))

metadata.12 <- relabun.lachno12 %>%
  select(sampleID, timepoint)

otu <- as.matrix(otu.12)
mode(otu) <- "numeric"
otu[is.na(otu)] <- 0
otu[otu < 0] <- 0

keep_samples <- rowSums(otu) > 0
otu <- otu[keep_samples, , drop = FALSE]
metadata.12 <- metadata.12 %>% filter(sampleID %in% rownames(otu))
otu <- otu[, colSums(otu) > 0, drop = FALSE]


```

#Compute distances + PCA for mimic 12
```{r}
dist_bray_12 <- vegdist(otu, method = "bray")
ord_12 <- cmdscale(dist_bray_12, k = 2, eig = TRUE)

ordination_df_12 <- data.frame(
  sampleID = rownames(ord_12$points),
  PC1 = ord_12$points[,1],
  PC2 = ord_12$points[,2]
) %>% left_join(metadata.12, by = "sampleID")
```

#plot for mimic 12
```{r}
ggplot(ordination_df_12, aes(PC1, PC2, color = timepoint)) +
  geom_point(size = 3, alpha = 0.8) +
  stat_ellipse(type = "t", level = 0.68, linewidth = 0.8) +
  theme_bw() +
  labs(title = "PCoA - Bray–Curtis (mimic 12)", x = "PCoA 1", y = "PCoA 2")
```

#split metadata vs taxa for mimic 1
```{r}
metadata.1 <- relabun.lachno1 %>%
  select(sampleID, timepoint, gavage.simple, treatment)

otu.1 <- relabun.lachno1 %>%
  select(-timepoint, -gavage.simple, -treatment, -sampleID) %>% 
  as.data.frame()

rownames(otu.1) <- relabun.lachno1$sampleID

#compute distances

dist_bray_1 <- vegdist(otu.1, method = "bray")

ord.1 <- cmdscale(dist_bray_1, k=2, eig = TRUE)
ordination_df.1 <- data.frame(
  sampleID = rownames(ord.1$points),
  PC1 = ord.1$points[,1],
  PC2 = ord.1$points[,2]
) %>%
  left_join(metadata.1, by = "sampleID")

#compute centroids 

centroids_1 <- ordination_df.1 %>%
  group_by(timepoint) %>%
  summarize(PC1_centroid = mean(PC1),
            PC2_centroid = mean(PC2))

ordination_df.1 <- ordination_df.1 %>%
  left_join(centroids_1, by = "timepoint") %>%
  mutate(dist_to_centroid = sqrt((PC1 - PC1_centroid)^2 + (PC2 - PC2_centroid)^2))
```

#plot for mimic 1
```{r}
ggplot(ordination_df.1, aes(PC1, PC2, color = timepoint)) +
  geom_segment(aes(xend = PC1_centroid, yend = PC2_centroid),
               alpha = 0.4, linewidth = 0.4) +
  geom_point(size = 3, alpha = 0.8) +
  stat_ellipse(type = "t", level = 0.68, linewidth = 0.8) +
  theme_bw() +
  labs(title = "PCoA - Bray–Curtis (mimic 1)",
       x = "PCoA 1", y = "PCoA 2")
```
```{r}
ggplot(ordination_df.1, aes(x = timepoint, y = dist_to_centroid, fill = timepoint)) +
  geom_boxplot(alpha = 0.6) +
  geom_jitter(width = 0.1, alpha = 0.7) +
  theme_bw() +
  labs(y = "Distance to centroid (Euclidean)",
       x = "Timepoint",
       title = "Group dispersion (mimic 1)")
```
#split metadata vs taxa for mimic 16
```{r}
metadata.16 <- relabun.lachno16 %>%
  select(sampleID, timepoint, gavage.simple, treatment)

otu.16 <- relabun.lachno16 %>%
  select(-timepoint, -gavage.simple, -treatment, -sampleID) %>% 
  as.data.frame()

rownames(otu.1) <- relabun.lachno1$sampleID

#compute distances

dist_bray_1 <- vegdist(otu.1, method = "bray")

ord.1 <- cmdscale(dist_bray_1, k=2, eig = TRUE)
ordination_df.1 <- data.frame(
  sampleID = rownames(ord.1$points),
  PC1 = ord.1$points[,1],
  PC2 = ord.1$points[,2]
) %>%
  left_join(metadata.1, by = "sampleID")

#compute centroids 

centroids_1 <- ordination_df.1 %>%
  group_by(timepoint) %>%
  summarize(PC1_centroid = mean(PC1),
            PC2_centroid = mean(PC2))

ordination_df.1 <- ordination_df.1 %>%
  left_join(centroids_1, by = "timepoint") %>%
  mutate(dist_to_centroid = sqrt((PC1 - PC1_centroid)^2 + (PC2 - PC2_centroid)^2))
```
```{r}

```

