---
title: "Lachno phylogeny enrichment"
author: "Rebecca Hoyd"
date: "2/6/2024"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(stringr)
library(ggnewscale)
library(taxize)
library(phytools)
library(phylobase)
```

# Load data

```{r}
modres <- readRDS("../data/BEWELL/glm_binomial_model.RData")
mics <- read.csv("../data/BEWELL/bewell_taxonomy_relAbun.csv", stringsAsFactors = F)

responders <- paste0("HONC60-", c(68,79,84,85))
```

# Define lachno groups

## All Lachnospira in our microbe results

```{r}
lachno.members <- unique(subset(mics$Taxonomy, grepl("f__Lachno", mics$Taxonomy)))

lachno.remtax <- str_remove(lachno.members,".*\\|")
lachno.taxids <- unique(subset(mics$TaxNum, grepl("f__Lachno", mics$Taxonomy)))
writeLines(lachno.taxids, "../data/BEWELL/lachnospiraceae-taxids.txt")
```

## Lachno associated with BRB

```{r}
brb.enriched <- modres %>%
  filter(estimate > 0 &
           term %in% lachno.remtax &
           p.value < 0.1)


```

## Lachno present in responders

```{r}
lachno.resp <- mics %>%
  filter(Subject.ID %in% responders & Taxonomy %in% lachno.members)
lachno.resp.tax <- unique(lachno.resp$Taxonomy)
lachno.resp.remtax <- gsub(".*\\|", "", lachno.resp.tax)

```

```{r}
brb.enriched.taxids <- mics %>%
  select(Taxonomy, TaxNum) %>%
  distinct() %>%
  mutate(nodeleaf = gsub(".*\\|", "", Taxonomy)) %>%
  filter(nodeleaf %in% brb.enriched$term)
```

# Format for phylo plot


```{r}
lachspecs <- gsub(".*(g__.*)", "\\1", subset(lachno.members, grepl("s__", lachno.members)))

lach.genspec <- as.data.frame(cbind(genus = gsub("\\|.*", "", lachspecs),
                                   species = gsub(".*\\|", "", lachspecs))) 

# Get the coordinates of the points
lach.genspec.xvals <- lach.genspec %>%
  arrange(genus) %>%
  mutate(specx = row_number(),
         specy = 4,
         geny = 3) %>%
  mutate(genx = median(specx), .by = genus)

lachno.pointsdf <- lach.genspec.xvals %>%
  select(genus, genx, geny) %>%
  bind_rows(select(lach.genspec.xvals, species, specx, specy) %>%
              rename("genus" = "species", "genx" = "specx", "geny" = "specy"))

# Add enriched info to the points
lachno.pointsdf.BRB <- lachno.pointsdf %>%
  mutate(colgroup = "Enriched in BRB",
         colcode = ifelse(genus %in% brb.enriched$term, "BRB", NA))
lachno.pointsdf.resp <- lachno.pointsdf %>%
  mutate(colgroup = "Enriched in responders",
         colcode = ifelse(genus %in% lachno.resp.remtax, "Responder", NA))

lachno.points.plot <- bind_rows(lachno.pointsdf.BRB, lachno.pointsdf.resp)

# Build tree structure

add.lachconnect <- lach.genspec.xvals %>%
  select(-species, -specx, -specy) %>%
  mutate(specx = genx, specy = 1) %>%
  distinct()
add.treebranch <- lach.genspec.xvals %>%
  mutate(x1 = min(specx),
         x2 = max(specx),
         .by = "genus") %>%
  select(genus, x1, x2) %>%
  distinct() %>%
  rename("specx" = "x1",
         "genx" = "x2") %>%
  mutate(genus = paste0(genus, "_branch"),
         specy = 3, geny = 3
         ) 
lachno.genspec.plot <- lach.genspec.xvals %>%
  mutate(genx = specx) %>%
  bind_rows(add.lachconnect) %>%
  bind_rows(add.treebranch)

# lach.treedf <- lach.genspec %>%
#   group_by(genus) %>%
#   summarise("leaves" = paste0("(", paste(species, collapse = ","), ")"))
# 
# lachno.tree <- paste0("(", paste(lach.treedf$leaves, collapse = ","), ");")
# lachno.tree2 <- read.tree(text = lachno.tree, file = NULL)
```
# Plot

```{r}
# plotTree(lachno.tree2)
```

```{r}
ggplot() +
  facet_wrap(vars(colgroup)) +
  geom_segment(data = lachno.genspec.plot, aes(x = genx, y = geny, xend = specx, yend = specy)) +
  geom_point(data = lachno.points.plot, aes(x = genx, y = geny, color = colcode),
             show.legend = F, size = 2.5) +
  coord_polar() +
  scale_color_brewer(palette = "Set1") +
  labs(x = "", y = "") +
  lims(y = c(1,4), x = c(0,50)) +
  theme_bw() +
  theme(axis.text = element_blank(),
        axis.ticks = element_blank(),
        panel.grid = element_blank())
ggsave("../figures/circle_lachno-brb-responder.png")
```

# Alternate coloring

## Get BRB fold change

```{r}
fc.prep <- mics %>%
  filter(Taxonomy %in% lachno.members & grepl("g__", Taxonomy)) %>%
  mutate(species = gsub(".*\\|", "", Taxonomy),
         genus = gsub(".*\\|(g__.*)", "\\1", Taxonomy),
         brb.code = ifelse(variable == "tB", "treatment", "control")) 

fc.spec <- fc.prep %>%
  filter(!grepl("g__", species)) %>%
  select(sample, species, RelAbun, brb.code) %>%
  pivot_wider(names_from = "species", values_from = "RelAbun", values_fill = 0) %>%
  pivot_longer(-c("sample", "brb.code"), names_to = "species", values_to = "RelAbun") %>%
  group_by(species, brb.code) %>%
  summarise(mean.ra = mean(RelAbun)) %>%
  pivot_wider(names_from = "brb.code", values_from = "mean.ra") %>%
  mutate(l2fc = log(treatment/control),
         l2fc = case_when(!l2fc %in% c(Inf, -Inf) ~ l2fc)) %>%
  rename("genus" = "species") %>%
  select(genus, l2fc)

fc.gen <- fc.prep %>%
  mutate(genus = str_remove(genus, "\\|s__.*")) %>%
  select(sample, genus, RelAbun, brb.code) %>%
  group_by(sample, genus, brb.code) %>%
  summarise(RelAbun = sum(RelAbun)) %>%
  ungroup() %>%
  pivot_wider(names_from = "genus", values_from = "RelAbun", values_fill = 0) %>%
  pivot_longer(-c("sample", "brb.code"), names_to = "genus", values_to = "RelAbun") %>%
  group_by(genus, brb.code) %>%
  summarise(mean.ra = mean(RelAbun)) %>%
  pivot_wider(names_from = "brb.code", values_from = "mean.ra") %>%
  mutate(l2fc = log(treatment/control),
         l2fc = case_when(!l2fc %in% c(Inf, -Inf) ~ l2fc)) %>%
  select(genus, l2fc)

points.foldchange <- bind_rows(fc.spec, fc.gen) %>%
  right_join(lachno.pointsdf.BRB)
```

## Get Responder median RAs

```{r}
respRA.prep <- mics %>%
  filter(Taxonomy %in% lachno.members & Subject.ID %in% responders & grepl("g__", Taxonomy)) %>%
  mutate(species = gsub(".*\\|", "", Taxonomy),
         genus = gsub(".*\\|(g__.*)", "\\1", Taxonomy)) 

respRA.spec <- respRA.prep %>%
  filter(!grepl("g__", species)) %>%
  select(sample, species, RelAbun) %>%
  pivot_wider(names_from = "species", values_from = "RelAbun", values_fill = 0) %>%
  pivot_longer(-sample, names_to = "species", values_to = "RelAbun") %>%
  summarise(spec.RA = median(RelAbun), .by = species) %>%
  rename("genus" = "species")

respRA.gen <- respRA.prep %>%
  mutate(genus = str_remove(genus, "\\|s__.*")) %>%
  group_by(sample, genus) %>%
  summarise(RelAbun = sum(RelAbun)) %>%
  ungroup() %>%
  pivot_wider(names_from = "genus", values_from = "RelAbun", values_fill = 0) %>%
  pivot_longer(-sample, names_to = "genus", values_to = "RelAbun") %>%
  summarise(spec.RA = median(RelAbun), .by = genus) 

points.respRA <- bind_rows(respRA.gen, respRA.spec) %>%
  right_join(lachno.pointsdf.resp) %>%
  drop_na(colcode)

```

## Complicated plot command

```{r}
ggplot() +
  facet_wrap(vars(colgroup)) +
  geom_segment(data = lachno.genspec.plot, aes(x = genx, y = geny, xend = specx, yend = specy)) +
  geom_point(data = points.foldchange, aes(x = genx, y = geny, color = l2fc),
             size = 2.5) +
  coord_polar() +
  scale_color_viridis_c(option = "A", name = "log2FoldChange") +
  labs(x = "", y = "") +
  lims(y = c(1,4), x = c(0,50)) +
  theme_bw() +
  theme(axis.text = element_blank(),
        axis.ticks = element_blank(),
        panel.grid = element_blank()) +
  new_scale_color() +
  geom_point(data = rename(points.respRA, "l2fc" = "spec.RA"), aes(x = genx, y = geny, color = l2fc),
             size = 2.5) +
  scale_color_continuous(type = "viridis", name = "Median RA\nin Responders")
ggsave("../figures/circle_lachno-brb-responder_2scale.png")
```

# Phylogenetic tree structure

## Retrieve new segments and point coords for plot

```{r}
lachno.taxids.leaf <- as.numeric(str_remove(subset(lachno.taxids, str_count(lachno.taxids, "\\|") == 6),
                                             ".*\\|"))
taxize_lachnos <- classification(lachno.taxids.leaf, db = "ncbi")
ncbitree <- class2tree(taxize_lachnos, check = T)
```

```{r}
png("../figures/phylogeny_default-lachno-plot.png")
plot(ncbitree)
dev.off()

plot(ncbitree, show.node = T)
```

need to do: take points df and get it joined to the older continuous variable tables

```{r}
getPhyloSegments <- function(phylotree, taxids){
  
  taxidkey <- as.data.frame(cbind(tname = phylotree$tip.label,
                                  taxid = as.character(taxids)))
  
  # Generate xy coords for edges and nodes
  tmp1 <- as(phylotree, "phylo4")
  tmp <- phylobase::phyloXXYY(tmp1)
  
  # Handle node labelling. To explain this strange set up: in the label vector, it presents the tip labels and then the internal labels, each in order. However, the xx and yy values resulting from phylobase XXYY have ALL nodes, internal and external, listed in order. It is possible to determine which nodes are tips as they have the maximum x value, so I separate the database via this measure and preserve its order so that I can apply each set of node lables and rejoin.
  nodepts.unlab <- as.data.frame(cbind(xx = tmp$xx, yy = tmp$yy)) %>%
    mutate(xx = as.numeric(xx),
           yy = as.numeric(yy))
  nodepts.unlab$tname <- tmp$phy@label[tmp$phy@edge[,2]]

  # nodepts.tip <- nodepts.unlab %>%
  #   filter(xx >= .9999)
  # nodepts.tip$tname <- tmp$phy@label[1:length(phylotree$tip.label)]

  # nodepts.int <- nodepts.unlab %>%
  #   filter(xx < .9999)
  # nodepts.int$tname <- tmp$phy@label[(length(phylotree$tip.label)+1):length(tmp$phy@label)]

  nodepts <- nodepts.unlab %>%
    left_join(taxidkey)
  
  
  # Make segments into a long df, instead of separated by horizontal and vertical segments.
  segs1 <- as.data.frame(tmp$segs)[,1:4]
  segs2 <- as.data.frame(tmp$segs)[,5:8]
  colnames(segs2) <- gsub("h", "v", colnames(segs2))
  segs <- bind_rows(segs1,segs2)
  
  # I still don't understand the node ordering, but I now believe I can make it work by pulling the node data from the edge data frame, where at least everything is labelled.
  nodelabs.alt <- as.data.frame(tmp$segs)
  nodelabs.alt$egde <- getEdge(tmp$phy)
  nodelabs.alt <- nodelabs.alt %>%
    separate(egde, into = c("ancestor", "descendent"))
  
  
  phylcoords <- list(segs, nodepts)
  names(phylcoords) <- c("Segments", "Nodes")
  return(phylcoords)
}

phylo.coords <- getPhyloSegments(ncbitree$phylo, lachno.taxids.leaf)

# phylo.coords$Nodes$taxid <- as.character(c(lachno.taxids.leaf, rep(NA, nrow(phylo.coords$Nodes) - length(lachno.taxids.leaf))))
```

```{r}
manualCoords <- function(phytree){
  edgepairs <- phytree$edges
  edgelengths <- phytree$edge.length
  
  allnodes <- c(phytree$tip.label, phytree$node.label)
  is_tip <- phytree$edge[,2] <= length(phytree$tip.label)
  ordered_tips <- tree2$edge[is_tip, 2]
  tree2$tip.label[ordered_tips]
}
```

## Matching to continuous data for points

Note on "missing" genera nodes: when a genera only has one member, taxize builds a tree omitting that genera and moving straight to the species. I believe this is appropriate for out data, so I am not going to try and force it to display a node. It also looks like that would be really difficult.
```{r}
gen.match <- as.data.frame(cbind(genus = subset(lachno.pointsdf$genus, grepl("g__", lachno.pointsdf$genus)))) %>%
  mutate(tname = str_remove(genus, "g__")) %>%
  distinct()
  # mutate(goodmatch = tname %in% phylo.coords$Nodes$tname)



# Species matching
pdf.matching <- as.data.frame(cbind(taxid = gsub(".*\\|", "", lachno.taxids),
                                    genus2 = lachno.remtax))

nodes.forjoin <- phylo.coords$Nodes %>%
  left_join(pdf.matching) %>%
  left_join(gen.match) %>%
  mutate(genus = coalesce(genus, genus2)) %>%
  drop_na(genus)
```

```{r}
phylopoints.BRB <- nodes.forjoin %>%
  left_join(points.foldchange) %>%
  distinct() %>%
  mutate(colgroup = gsub("in", "by", colgroup))

phylopoints.resp <- nodes.forjoin %>%
  left_join(points.respRA) %>%
  distinct() %>%
  drop_na(colcode) %>%
  mutate(colgroup = gsub("Enriched", "Abundance", colgroup))
```

```{r}
ggplot(phylo.coords$Segments) +
  geom_segment(aes(x = v0x, y = v0y, xend = v1x, yend = v1y)) +
  geom_point(data = as.data.frame(cbind(xx = tmp$xx, yy = tmp$yy)), aes(y = yy, x = xx), inherit.aes = F)
```

## Plot

```{r}
ggplot(phylo.coords$Segments) +
  facet_wrap(vars(fct_relevel(colgroup, "Enriched by BRB"))) +
  geom_segment(aes(y = v0x, x = v0y, yend = v1x, xend = v1y)) +
  geom_point(data = phylopoints.BRB, aes(x = yy, y = xx, color = l2fc),
             size = 2.5) +
  coord_polar() +
  scale_color_viridis_c(option = "A", name = "log2FoldChange") +
  lims(x = c(0,1.05)) +
  labs(x = "", y = "") +
  theme_bw() +
  theme(strip.text = element_text(size = 12),
        legend.text = element_text(size = 12),
        axis.text = element_blank(),
        axis.ticks = element_blank(),
        panel.grid = element_blank()) +
  new_scale_color() +
  geom_point(data = rename(phylopoints.resp, "l2fc" = "spec.RA"), aes(x = yy, y = xx, color = l2fc),
             size = 2.5) +
  scale_color_continuous(type = "viridis", name = "Median RA\nin Responders")
ggsave("../figures/circle_lachno_phylo-continuous.png")
```
