---
title: "Phylogenetic relatedness between B.mas, B.ob, A.rectal"
author: "Shiva Jahanbakhshi"
date: "2026-07-15"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## Load packages and GTDB files

Figure 5A uses the bacterial reference tree from [GTDB release 232](https://data.gtdb.ecogenomic.org/releases/release232/232.0/) (15 April 2026). Those files are not in this repository. The metadata file is 276 MB, which is over GitHub's file-size limit. Download the release 232 files and save them in `exploratory/data/` under the names this notebook reads:

- [bac120_r232.tree.gz](https://data.gtdb.ecogenomic.org/releases/release232/232.0/bac120_r232.tree.gz) as `bac120.tree.gz` (2.3 MB)
- [bac120_taxonomy_r232.tsv.gz](https://data.gtdb.ecogenomic.org/releases/release232/232.0/bac120_taxonomy_r232.tsv.gz) as `bac120_taxonomy.tsv.gz` (9.4 MB)
- [bac120_metadata_r232.tsv.gz](https://data.gtdb.ecogenomic.org/releases/release232/232.0/bac120_metadata_r232.tsv.gz) as `bac120_metadata.tsv.gz` (276 MB)

The three genomes marked on the tree, `RS_GCF_025147765.1` (*B. obeum*), `RS_GCF_000020605.1` (*A. rectalis*), and `RS_GCF_020687435.1` (*B. massiliensis*), are tips in this release. `/releases/latest/` also serves these files under the unversioned names, and it is release 232 as of September 2026. That directory will change when GTDB publishes the next release, so use the release 232 links above.

```{r}
library(tidyverse)
library(readr)
library(ape)
library(stringr)

# GTDB release 232. See the download links in the section above.
metadata <- read_tsv("../data/bac120_metadata.tsv.gz")
taxonomy <- read_tsv(
  "../data/bac120_taxonomy.tsv.gz",
  col_names = FALSE
)

tree <- read.tree("../data/bac120.tree.gz")

```

## Extract GTDB taxonomy levels
```{r}
metadata <- metadata %>%
  mutate(
    gtdb_family = str_extract(
      gtdb_taxonomy,
      "f__[^;]+"
    ),
    gtdb_genus = str_extract(
      gtdb_taxonomy,
      "g__[^;]+"
    ),
    gtdb_species = str_extract(
      gtdb_taxonomy,
      "s__[^;]+$"
    )
  )
```

## Find the GTDB names of the three target species
```{r}
target_metadata <- metadata %>%
  filter(
    str_detect(
      ncbi_organism_name,
      regex(
        paste(
          "Blautia massiliensis",
          "Blautia obeum",
          "Agathobacter rectalis",
          sep = "|"
        ),
        ignore_case = TRUE
      )
    )
  ) %>%
  select(
    accession,
    ncbi_organism_name,
    ncbi_strain_identifiers,
    gtdb_species,
    gtdb_representative
  )

target_metadata

#Display the unique NCBI-to-GTDB mappings

target_metadata %>%
  select(
    ncbi_organism_name,
    gtdb_species
  ) %>%
  distinct()

```
```{r}
## Define the GTDB species names of the three targets

target_species <- target_metadata %>%
  filter(
    gtdb_representative %in% c(TRUE, "t", "TRUE")
  ) %>%
  pull(gtdb_species) %>%
  unique()

target_species
```


## Find the representative genomes
```{r}
target_genomes <- metadata %>%
  filter(
    gtdb_species %in% target_species,
    gtdb_representative %in% c(TRUE, "t", "TRUE")
  ) %>%
  select(
    accession,
    ncbi_organism_name,
    gtdb_species
  )

target_genomes 

#Define the selected accessions in a consistent order

target_tips <- c(
  "RS_GCF_025147765.1",  # Blautia obeum
  "RS_GCF_000020605.1",  # Agathobacter rectalis
  "RS_GCF_020687435.1"   # Blautia massiliensis
)

target_names <- c(
  "B. obeum",
  "A. rectalis",
  "B. massiliensis"
)

target_colors <- c(
  "#009E73",
  "#0072B2",
  "#E69F00"
)

#Confirm that all three genomes are in the GTDB tree

target_tips %in% tree$tip.label
```
#all representative Lachnospiraceae genomes
## Extract all representative genomes in the family

```{r}
# Get all representative Lachnospiraceae genomes
lachno_df <- metadata %>%
  filter(
    gtdb_family == "f__Lachnospiraceae",
    gtdb_representative %in% c("t", "TRUE", TRUE)
  )

# Keep only accessions that also occur in the tree
lachno_tips <- intersect(
  lachno_df$accession,
  tree$tip.label
)

length(lachno_tips)
head(lachno_tips)

#prune
lachno_tree <- keep.tip(
  tree,
  lachno_tips
)

length(lachno_tree$tip.label)

# keep the target tips with the prefixes
target_tips <- c(
  "RS_GCF_025147765.1",  # Blautia obeum
  "RS_GCF_000020605.1",  # Agathobacter rectalis
  "RS_GCF_020687435.1"   # Blautia massiliensis / GTDB B. fusiformis
)

target_tips %in% tree$tip.label
```
## Find the locations of the three targets
```{r}
target_positions_all <- match(
  target_tips,
  lachno_tree$tip.label
)

target_positions_all
```

## Plot all Lachnospiraceae representatives
```{r}
pdf(
  "Lachnospiraceae_all_representatives.pdf",
  width = 11,
  height = 11
)

par(mar = rep(0.5, 4))

plot(
  lachno_tree,
  type = "fan",
  show.tip.label = FALSE,
  show.node.label = FALSE,
  edge.width = 0.6,
  no.margin = TRUE
)

tiplabels(
  tip = target_positions_all,
  pch = 21,
  bg = target_colors,
  col = "black",
  cex = 1.5
)

tiplabels(
  tip = target_positions_all,
  text = target_names,
  frame = "none",
  col = target_colors,
  font = 3,
  cex = 0.85,
  adj = c(-0.05, 0.5)
)

dev.off()
```
```{r}
genus_representatives <- metadata %>%
  filter(
    gtdb_family == "f__Lachnospiraceae",
    gtdb_representative == "t",
    !is.na(gtdb_genus),
    gtdb_genus != "g__"
  ) %>%
  arrange(
    gtdb_genus,
    accession
  ) %>%
  group_by(gtdb_genus) %>%
  slice(1) %>%
  ungroup() %>%
  pull(accession) %>%
  intersect(tree$tip.label)

tips_to_keep <- unique(
  c(
    genus_representatives,
    target_tips
  )
)

lachno_genus_tree <- keep.tip(
  tree,
  tips_to_keep
)

length(lachno_genus_tree$tip.label)
```
```{r}
target_positions_genus <- match(
  target_tips,
  lachno_genus_tree$tip.label
)

target_positions_genus
```
```{r}
pdf(
  "Lachnospiraceae_genus_representative_tree.pdf",
  width = 12,
  height = 12
)

par(mar = rep(1, 4))

plot(
  lachno_genus_tree,
  type = "fan",
  show.tip.label = FALSE,
  show.node.label = FALSE,
  use.edge.length = FALSE,
  edge.width = 0.7,
  no.margin = TRUE
)

tiplabels(
  tip = target_positions_genus,
  pch = 21,
  bg = target_colors,
  col = "black",
  cex = 1.7
)

tiplabels(
  tip = target_positions_genus,
  text = target_names,
  frame = "none",
  col = target_colors,
  font = 3,
  cex = 1,
  adj = c(-0.4, 0.5)
)

dev.off()
```

```{r fig5A}
# Rectangular tree for submitted Figure 5A. The fan PDFs above are the working layouts.
ordered <- reorder(ladderize(lachno_genus_tree), "cladewise")
edge <- ordered$edge
tip_y <- numeric(Ntip(ordered) + Nnode(ordered))
tips <- edge[, 2][edge[, 2] <= Ntip(ordered)]
tip_y[tips] <- seq_along(tips)
children <- split(edge[, 2], edge[, 1])
for (parent in rev(unique(edge[, 1]))) {
  kids <- children[[as.character(parent)]]
  tip_y[parent] <- mean(tip_y[kids])
}
node_x <- node.depth.edgelength(ordered)
horiz <- data.frame(
  x = node_x[edge[, 1]],
  xend = node_x[edge[, 2]],
  y = tip_y[edge[, 2]],
  yend = tip_y[edge[, 2]]
)
vert <- bind_rows(lapply(names(children), function(parent) {
  kids <- children[[parent]]
  data.frame(
    x = node_x[as.integer(parent)],
    xend = node_x[as.integer(parent)],
    y = min(tip_y[kids]),
    yend = max(tip_y[kids])
  )
}))
tip_lab <- data.frame(
  x = node_x[seq_len(Ntip(ordered))],
  y = tip_y[seq_len(Ntip(ordered))],
  accession = ordered$tip.label
)
tip_lab$label <- target_names[match(tip_lab$accession, target_tips)]
tip_lab$label[is.na(tip_lab$label)] <- ""
tip_lab$color <- target_colors[match(tip_lab$label, target_names)]
tip_lab$color[is.na(tip_lab$color)] <- "grey30"

fig5A <- ggplot() +
  geom_segment(data = horiz, aes(x = x, xend = xend, y = y, yend = yend)) +
  geom_segment(data = vert, aes(x = x, xend = xend, y = y, yend = yend)) +
  geom_text(data = tip_lab, aes(x = x, y = y, label = label, color = color),
            hjust = 0, nudge_x = 0.01, size = 2.5, fontface = "italic", show.legend = FALSE) +
  scale_color_identity() +
  theme_void()

ggsave("../../manuscript/figures/panels/fig5A_lachnospiraceae-tree.svg",
       fig5A,
       height = 7, width = 7)
```

